Fractional Types

### 2013-02-23 frac.tex
collapsed section 4.5 into 4.4

%%%%%%%%%%%%
\subsection{Feedback, Iteration, and Trace Operators}

@@

\noindent  The computation with {{true}} gives the same result.  This
implies that although our semantics is formally reversible, it 
does not result in isomorphisms.
Nevertheless, the next subsection illustrates why this is still useful.

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{(Finite) Relational Programming}
\label{ch3:sec:lp}

Relational programming leverages set-theoretic relations and their
composition to express computational tasks in a declarative way. 
% For instance, consider this example:
% {{parent = {[ (A,B), (B,C), (A,D) ]} }}
% {{grandparent = parent (o) parent}}
% \noindent The example defines a relation {{parent}} specified using a
% set of tuples and another relation {{grandparent}} specified using the
% composition of two parent relations. If we wrote this example in a
% relational language (e.g., Prolog) and we executed the query
% {{grandparent(A)}}, we would get the answer {{ {[C]} }}.
It turns out that with the addition of the multiplicative trace, and
the move to relations motivated in the previous section, we can
express relational programming.

Consider the relation {{R}} on booleans given by:

{{ {[(false,false), (false,true), (true,false) ]}. }}

\noindent We can define a combinator {{c_R}} whose denotation is {{R}},
by defining a combinator {{cInner : (a * bool) <-> (a * bool)}}
for some type {{a}} such that {{c_R = trace* cInner}}:

% %subcode{proof} include main
% %@ ~
% %@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
% %----
% %@ ~
% %@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
% %----
% %@ ~
% %@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
% %----
% %@ ~
% %@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
% %----
% %@ ~
% %@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
% %----
% %@ ~
% %@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
% %----
% %@ ~
% %@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
% %----
% %@ ~
% %@@ ((true,true),true) ~~cInner~~ ((false,false),true) 
% In the first three lines, the first argument is a fixed point and hence the
% multiplicative trace would map the second input to the second output
% producing the desired relation. In the remaining five cases, the first
% argument is not a fixed point and hence all these cases would be rejected as
% solutions to the constraint imposed by the multiplicative trace. It simply
% remains to find the actual realization of {{cInner}} that would produce the
% behavior above.

%subcode{opsem} include main
%! columnStyle = rcl
% cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;)  toffoli (;)
%        &&  ((swap* (;) cnot (;) swap*) (*) id) (;) toffoli (;)
%        &&  (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;) toffoli (;) (cnot (*) id))

\noindent where {{toffoli}} and {{cnot}} were defined for {{Pi}} in
\cite{rc2011}.
% %     
% % c_R &:& bool <-> bool
% % c_R &=& trace* cInner

>>>>

%%%%%%%%%%%%
\subsection{Trace Operators and (Finite) Relational Programming}

@@

%% XXX: do we want to cite something for relational programming?
\noindent  The computation with {{true}} gives the same result.  This
implies that although our semantics is formally reversible, it 
does not result in isomorphisms.
Nevertheless, this can still be useful, as it gives us access to
\emph{relational programming},
which uses set-theoretic relations and their
composition to express computational tasks in a declarative way. Their encoding
in {{PiTF}} uses fractional types (and specifically, their ability to encode the
multiplicative trace operator).
%% XXX: this falls kind of flat without some kind of example. do we want to
%% release code with the paper and mention it here?

% For instance, consider this example:
% {{parent = {[ (A,B), (B,C), (A,D) ]} }}
% {{grandparent = parent (o) parent}}
% \noindent The example defines a relation {{parent}} specified using a
% set of tuples and another relation {{grandparent}} specified using the
% composition of two parent relations. If we wrote this example in a
% relational language (e.g., Prolog) and we executed the query
% {{grandparent(A)}}, we would get the answer {{ {[C]} }}.

% Consider the relation {{R}} on booleans given by:
% 
% {{ {[(false,false), (false,true), (true,false) ]}. }}
% 
% \noindent We can define a combinator {{c_R}} whose denotation is {{R}},
% by defining a combinator {{cInner : (a * bool) <-> (a * bool)}}
% for some type {{a}} such that {{c_R = trace* cInner}}:
% 
% % %subcode{proof} include main
% % %@ ~
% % %@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
% % %----
% % %@ ~
% % %@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
% % %----
% % %@ ~
% % %@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
% % %----
% % %@ ~
% % %@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
% % %----
% % %@ ~
% % %@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
% % %----
% % %@ ~
% % %@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
% % %----
% % %@ ~
% % %@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
% % %----
% % %@ ~
% % %@@ ((true,true),true) ~~cInner~~ ((false,false),true) 
% % In the first three lines, the first argument is a fixed point and hence the
% % multiplicative trace would map the second input to the second output
% % producing the desired relation. In the remaining five cases, the first
% % argument is not a fixed point and hence all these cases would be rejected as
% % solutions to the constraint imposed by the multiplicative trace. It simply
% % remains to find the actual realization of {{cInner}} that would produce the
% % behavior above. 
% 
% %subcode{opsem} include main
% %! columnStyle = rcl
% % cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
% % cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;)  toffoli (;)
% %        &&  ((swap* (;) cnot (;) swap*) (*) id) (;) toffoli (;)
% %        &&  (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;) toffoli (;) (cnot (*) id))
% 
% \noindent where {{toffoli}} and {{cnot}} were defined for {{Pi}} in
% \cite{rc2011}.
% %     
% % c_R &:& bool <-> bool
% % c_R &=& trace* cInner
